A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-28 manuscript.qmd
update mlpwr reference and add footnote on performing cost-efficient design analysis with mlpwr; fix some typos

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. In @tbl-rpackages, we review available software packages that can be used to perform power analysis for multilevel models in `r cite_r()`. Power analysis methods can be categorized into formula-based, summary-statistics-based and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on exact formulas to calculate power directly. Summary-statistics-based methods use statistical theory to approximate power based on formula-based methods developed for simple t-tests. Simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically, that is what percentage of simulated datasets produces a significant p-value.
Available formula-based and summary-statistics-based software packages for multilevel models often do not include GLMMs or are limited to simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a]. However, many of these tutorials focus on LMMs and the most common study designs [see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers using more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions about the model structure that should align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021). We will discuss a number of strategies on how to specify model parameters for application-specific, tailored data simulations in a later chapter.

@@

```{r}
#| label: fig-margdist1
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions."
#| echo: true
library(ggdist)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
@fig-margdist1 shows the model-implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered previously. This difference increases with the variability of the random effects.
We can use plots like @fig-margdist1 as a useful tool to decide whether the specified standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$) are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

@@

When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other based on practical considerations. Although it is recommended to justify the power [@lakensSampleSizeJustification2022], we adopt the common heuristic to target a power of 0.8 to detect an effect of the expected size implied by our data simulation. This could be achieved by collecting data from 200 subjects (about 25% of which will be experts), each completing the same 50 items (with advice present in about 67% of cases, which is correct in about 80% of cases with present advice). If collecting data from 200 subjects is not feasible, an alternative would be to recruit 150 subjects but increase the length of the experiment to over 70 items. However, 70 items might take too long to complete for the radiologists participating in the study, who have a busy schedule. The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items if it is feasible to recruit an even higher number of subjects (\> 250, to be determined by additional rounds of power analysis). Design parameters that also affect power, and which could be investigated in the simulation to find a more optimal trade-off, are the ratio of experts to students, the frequency of whether advice is presented at all, and whether it is correct or incorrect.

@@

2. The ability to conduct simulation-based sample size planning becomes an increasingly valuable skill that should be taught to experimental researchers. By incorporating such training into research methods courses and workshops, researchers can improve the quality of their experimental designs and enhance the rigor of their studies. The need to reason about how to simulate plausible data that is in line with the research hypothesis, while not violating domain expertise on how plausible data should look might also contribute to planning more insightful studies that can answer more precise research questions [@yarkoniGeneralizabilityCrisis2022].

>>>>

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. In @tbl-rpackages, we review available software packages that can be used to perform power analysis for multilevel models in `r cite_r()`. Power analysis methods can be categorized into formula-based, summary-statistics-based and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on exact formulas to calculate power directly. Summary-statistics-based methods use statistical theory to approximate power based on formula-based methods developed for simple t-tests. Simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically, that is what percentage of simulated datasets produces a significant p-value.
Available formula-based and summary-statistics-based software packages for multilevel models often do not include GLMMs or are limited to simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2023; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a]. However, many of these tutorials focus on LMMs and the most common study designs [see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers using more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions about the model structure that should align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021). We will discuss a number of strategies on how to specify model parameters for application-specific, tailored data simulations in a later chapter.

@@

```{r}
#| label: fig-margdist1
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions."
#| echo: true
library(ggdist)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"student, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"student, incorrect advice" = "0:1:0", "expert, incorrect advice" = "1:1:0",
"student, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
@fig-margdist1 shows the model-implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered previously. This difference increases with the variability of the random effects.
We can use plots like @fig-margdist1 as a useful tool to determine whether the specified standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$) are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

@@

When target power has been specified, the number of subjects and the number of items in our study design can be traded against each other based on practical considerations. Although it is recommended to justify the power [@lakensSampleSizeJustification2022], we adopt the common heuristic to target a power of 0.8 to detect an effect of the expected size implied by our data simulation. This could be achieved by collecting data from 200 subjects (about 25% of which will be experts), each completing the same 50 items (with advice present in about 67% of cases, which is correct in about 80% of cases with present advice). If collecting data from 200 subjects is not feasible, an alternative would be to recruit 150 subjects but increase the length of the experiment to over 70 items. However, 70 items might take too long to complete for the radiologists participating in the study, who have a busy schedule. The simulation suggests that it might also be possible to plan a shorter experiment with only 30 items if it is feasible to recruit an even higher number of subjects (\> 250, to be determined by additional rounds of power analysis). Design parameters that also affect power, and which could be investigated in the simulation to find a more optimal trade-off, are the ratio of experts to students, the frequency of whether advice is presented at all, and whether it is correct or incorrect.^[An advanced alternative to our heuristic trade-off between subjects and items would be to explicitly optimize for cost-efficient study designs using the *mlpwr* package [@zimmerSampleSizePlanning2023]. The package performs simulation based power analysis in a surrogate modeling framework that can evaluate power for a large grid of design combinations more efficiently. The framework also requires users to write their own `sim_and_analyse` function and can be considered a great extension on what we cover in our tutorial.]

@@

2. The ability to conduct simulation-based sample size planning becomes an increasingly valuable skill that should be taught to experimental researchers. By incorporating such training into research methods courses and workshops, researchers can improve the quality of their experimental designs and enhance the rigor of their studies. The need to reason about how to simulate plausible data that is in line with the research hypothesis, while not violating domain expertise on how plausible data should look, might also contribute to planning more insightful studies that can answer more precise research questions [@yarkoniGeneralizabilityCrisis2022].
